---
title: "Performance of deep learning methods and open satellite time series in detecting low-severity spruce beetle mortality in Southcentral Alaska"
date: today
author:
  - name: Seamus Murphy
    orcid: 0000-0002-1792-0351
abstract: |
  Placeholder. The confirmatory analysis is gated behind pre-registration; this draft
  carries the framing, the data provenance and the planned design only. See
  `docs/science-superpowers/questions/2026-08-10-detectability-floor.md`.
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
    code-fold: false
    df-print: kable
  docx:
    prefer-html: true
    reference-doc: ../04.references/style-formal.docx
    highlight-style: pygments
execute:
  eval: true
  echo: true
  warning: false
  message: false
  error: true
  comment: NA
engine: knitr
citeproc: true
editor: visual
bibliography: ../04.references/references.bib
csl: ../04.references/apa.csl
---

```{r}
#| label: setup
#| echo: true
#| output: false
library(sf)
library(terra)
root <- normalizePath("..")
inputs <- file.path(root, "02.inputs")
```

# Introduction

Spruce beetle (*Dendroctonus rufipennis*) has killed spruce across more than a million
hectares of Southcentral Alaska since the outbreak began around 2016, yet the region's
operational damage record rests on aerial sketch mapping with known gaps in low-severity
and dispersed mortality [@Zwieback_2024]. Mapping that mortality from high-resolution
commercial imagery is feasible: a convolutional network on WorldView-2 scenes identified
dead spruce at 95% average accuracy and showed that roughly 90% of killed trees sit in
stands whose areal severity is below 0.05, beneath the change signal that Landsat SWIR
reflectance carries, which relates clearly to severity only above 0.1 [@Zwieback_2024].
Whether freely available satellite time series can reach below that floor is untested.
Temporal deep learning on Landsat subsequences recovers gradual and partial boreal
disturbances that index-based change detection misses, but the evidence comes from
defoliator outbreaks in eastern Canada [@Perbet_2024]. Reviews of the wider field flag
insect damage attribution and low-severity detection as priority gaps in western North
America, with Alaska excluded from their scope [@Hicke_2024], and show that early
detection in the intensively studied European *Ips typographus* system remains below
operational accuracy [@Kautz_2024], though the multi-year green stage of *Dendroctonus*
systems makes the timing problem less punishing here. Prior Alaska-wide work attributed
insect damage only through coarse survey perimeters intersected with Landsat trends
[@Pastick_2018] or modeled beetle occurrence rather than damage [@Zabihi_2021], and
individual-tree mortality benchmarks exist only outside Alaska [@Khatri_Chhetri_2024].

**Research question.** Down to what stand-scale areal severity can models trained on
freely available Sentinel-2 and Landsat time series detect spruce beetle associated
mortality in Southcentral Alaska, and does temporal deep learning lower that floor
relative to SWIR change detection?

::: {.callout-note}
## Pipeline status
This draft precedes pre-registration. Under the project's confirmatory-analysis gate, no
outcome data have been analysed: the live code below is restricted to provenance
verification and descriptive context, and the planned-analysis chunks are stubs with
`eval: false`. They activate only after the pre-registration document is complete.
:::

# Data

All inputs are open-source; retrieval is scripted in `05.scripts/00_download_inputs.sh`
and provenance recorded in `02.inputs/README.md`.

## Integrity of retrieved files

Every file analysed below is checksum-verified at render time against the manifest
recorded at download (2026-08-10).

```{r}
#| label: checksums
manifest <- read.table(file.path(inputs, "SHA256SUMS.txt"),
                       col.names = c("sha256", "file"))
manifest$verified <- vapply(manifest$file, function(f) {
  h <- system2("shasum", c("-a", "256", shQuote(file.path(inputs, f))),
               stdout = TRUE)
  identical(strsplit(h, " ")[[1]][1],
            manifest$sha256[manifest$file == f])
}, logical(1))
stopifnot(all(manifest$verified))
manifest[, c("file", "verified")]
```

## Reference severity rasters (Zwieback et al. 2024)

Stand-scale areal severity of beetle-associated dead spruce for the Talkeetna to Byers
Lake area, 2021, derived by @Zwieback_2024 from a convolutional network applied to
WorldView-2 imagery and released at three stand scales (Zenodo 10.5281/zenodo.8423568,
licence clarification pending; validation use only). Metadata inspected here; values are
not summarised or plotted before pre-registration.

```{r}
#| label: severity-metadata
sev <- lapply(c("merged_ref_h_c30c.tif", "merged_ref_h_c90c.tif",
                "merged_ref_h_c250c.tif"),
              function(f) rast(file.path(inputs, "zwieback-severity", f)))
data.frame(
  file       = vapply(sev, function(r) basename(sources(r)), ""),
  resolution = vapply(sev, function(r) paste(round(res(r)), collapse = " x "), ""),
  ncell      = vapply(sev, ncell, numeric(1)),
  crs        = vapply(sev, function(r) st_crs(crs(r))$Name, "")
)
```

## Crown-scale labels and pre-labelling model (CC-BY 4.0)

The hand-delineated dead-spruce crown labels and trained network of @Zwieback_2024
(Zenodo 10.5281/zenodo.10569990 and 10.5281/zenodo.10569975) serve as the starting point
for our own crown digitisation on the State of Alaska Geoportal Maxar mosaics (2020 at
50 cm, 2023 at 30 cm; streamed under the state's university/non-commercial terms).

```{r}
#| label: labels-inventory
lab_dir <- file.path(inputs, "zwieback-labels", "gpkg")
labs <- list.files(lab_dir, pattern = "_labels\\.gpkg$")
exts <- list.files(lab_dir, pattern = "_extent\\.gpkg$")
c(label_areas = length(labs), extent_files = length(exts))
```

## Aerial detection survey polygons (USFS IDS, Region 10)

Sketch-mapped damage polygons for Alaska, all survey years, US public domain. These are
weak labels: positional accuracy is coarse, low-severity damage is under-recorded, and
the intensity attribute counts faded trees per treed area, which is not commensurable
with pixel-fraction severity [@Zwieback_2024]. Annual spruce beetle extent is tabulated
here as published descriptive context, not as an outcome.

```{r}
#| label: ids-context
gdb <- file.path(inputs, "ids-region10", "AK_Region10_AllYears.gdb")
st_layers(gdb)
# The OGR SQL dialect has no GROUP BY; aggregate through GDAL's SQLite dialect
# instead, which also avoids loading 151k polygon geometries on this machine.
ids_sql <- function(sql) {
  tmp <- tempfile(fileext = ".csv")
  gdal_utils("vectortranslate", gdb, tmp,
             options = c("-dialect", "SQLite", "-sql", sql))
  read.csv(tmp)
}
sb <- ids_sql('
  SELECT SURVEY_YEAR, DCA_CODE, DCA_COMMON_NAME,
         COUNT(*) AS n_polygons, SUM(ACRES) AS acres
  FROM DAMAGE_AREAS_FLAT_AllYears_AK_Rgn10
  WHERE DCA_COMMON_NAME LIKE "%spruce beetle%"
  GROUP BY SURVEY_YEAR, DCA_CODE')
aggregate(cbind(n_polygons = sb$n_polygons,
                hectares = round(sb$acres * 0.40468564)),
          by = list(agent = sb$DCA_COMMON_NAME), FUN = sum)
data.frame(year = sb$SURVEY_YEAR[sb$DCA_CODE == 11009],
           n_polygons = sb$n_polygons[sb$DCA_CODE == 11009],
           hectares = round(sb$acres[sb$DCA_CODE == 11009] * 0.40468564))
```

## Predictor imagery (not yet touched)

Sentinel-2 (2016 to present) and Landsat surface reflectance will be assembled in Google
Earth Engine at the analysis stage; the IFSAR 5 m DEM and the Macander conifer-fraction
map enter as covariates. None of these are loaded before pre-registration.

# Planned methods

The design document precedes this section and will be superseded by the
pre-registration; the chunks below are inert stubs that fix the intended shape of the
analysis.

1. **Labels.** Dead-spruce crowns digitised on the 2020 and 2023 Geoportal Maxar
   mosaics over stratified sample blocks, pre-labelled with the released network of
   @Zwieback_2024 and hand-corrected, then aggregated to stand-cell severity fractions
   at 90 m (primary) and 30 m (secondary).
2. **Models.** A temporal deep learning model on Sentinel-2/Landsat subsequences
   (TempCNN and Transformer, following @Perbet_2024) against a pre- to post-outbreak
   SWIR reflectance change baseline (the null model, following @Zwieback_2024).
3. **Evaluation.** Block spatial cross-validation; user's and producer's accuracy and
   severity regression error reported per severity stratum; the detectability floor is
   the lowest stratum at which skill exceeds the pre-registered threshold. The
   independent Zwieback severity raster validates in the TBL area.

```{r}
#| label: planned-severity-strata
#| eval: false
# Activated after pre-registration. Severity strata for skill-by-severity reporting.
strata <- c(0, 0.01, 0.03, 0.05, 0.10, 0.20, 1)
```

```{r}
#| label: planned-baseline
#| eval: false
# Activated after pre-registration. SWIR change baseline: pre (2014-2016) minus
# post (2019-2021) summer means per stand cell, per Zwieback et al. (2024).
```

# Results

Gated: no results before pre-registration is complete.

# Data availability

All inputs are publicly retrievable; the download script, checksums and provenance
manifest are in the project repository. The Zwieback severity rasters carry no declared
licence and are used here for validation only pending clarification with the author.

# References
